Visual diagnostics for constrained optimisation with application to guided tours

### 2020-12-10 paper/paper.Rmd
optim method section is good for proofreading

Visualisation is widely used in exploratory data analysis [@tukey1977exploratory; @unwin2015graphical;  @healy2018data; @wilke2019fundamentals]. Presenting information in graphics often unveils information that would otherwise not be aware of and provides a more comprehensive understanding of the problem at hand. The work presented in this paper brings visualization tools into optimisation problems with an aim to better understand the performance of the optimisers in practice. This motivates our work of creating plots to diagnose optimisation algorithms which in turn can be also used not only to understand optimization performance but also to understand and compare the behaviour of different algorithms.

@@

The motivation of this paper is the optimisation problem arising in the projection pursuit guided tour [@buja2005computational] which is an exploratory data analysis tool that is defined to detect *interesting structures* or features  in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours [@cook2008grand]. The target of the optimisation is to identify the most *interesting* low-dimensional views of the data given by a corresponding projection matrix. The most *interesting* structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space [@cook1993projection].

@@

Optimization methods can be divided in various classes (NOTE: do we want one more sentence on various class: global optim, convex optim, stochastic optim etc?) [@kelley1999iterative; @boyd2004convex; @fletcher2013practical]. In this paper, we focused on methods that rely on the objective function derivatives known as gradient methods, and those which are known as derivative free methods where optimization does not rely on the gradient or derivatives information.

In gradient descent methods the goal is to find the minimum of an objective function while in gradient ascent methods the interest is in the maximum. The minimum is searched for by taken steps  proportional to the negative gradient or of the function at the evaluation point (analogously positive gradient for gradient ascent methods), that is $x_{k+1} = x_k -\alpha_kS_k$ where $\alpha_k$ is the step size or the learning rate [@fletcher2013practical].  The step size is the key in the success of the optimization problem as inadequate step sizes might lead to a failure in convergence. These methods are known to be relative slow and can have convergence issues when problems are poorly conditioned [@trefethen1997numerical]. In order to solve some of those problems, line minimization or line search methods were proposed [@shi2004convergence] where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate $x_{1}$, a search direction $S_k$ for each iteration where the objective is to find the value of $\alpha_k \in \mathbb{R}$ that minimizes $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ (this step is called the line search). Then move on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and repeat the process until the desire convergence is reached.
Modern development of line search methods focuses on improving the computation of the searching direction $S_{k}$, and on approximations of the step size $\alpha_k$ together with reducing the computational storing
burden of the intermediate computations while catering for practical optimisation problems. In the next section two line search methods for the optimisation used in the projection pursuit guided tour will be presented.

>>>>

Visualisation is widely used in exploratory data analysis [@tukey1977exploratory; @unwin2015graphical;  @healy2018data; @wilke2019fundamentals]. Presenting information in graphics often unveils information that would otherwise not be aware of and provides a more comprehensive understanding of the problem at hand. The work presented in this paper brings visualization tools into optimisation problems with an aim to better understand the performance of the optimisers in practice.

@@

The focus of this paper is on the optimisation problem arising in the projection pursuit guided tour [@buja2005computational] which is an exploratory data analysis tool that is defined to detect *interesting structures* or features in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours [@cook2008grand]. The target of the optimisation is to identify the most *interesting* low-dimensional views of the data given by a corresponding projection matrix. The most *interesting* structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space [@cook1993projection].

@@

Optimization methods can be divided in various classes, such as global optimisation [@kelley1999iterative; @fletcher2013practical], convex optimisation [@boyd2004convex], stochastic optimisation [@nocedal2006numerical] . In this paper, we focused on methods that rely on the objective function derivatives known as gradient methods, and those which are known as derivative free methods where optimization does not rely on the gradient or derivatives information.

In gradient ascent methods the goal is to find the maximum of an objective function while in gradient descent methods the interest is in the minimum. The maximum is searched for by taken steps  proportional to the gradient of the function at the evaluation point that is $x_{k+1} = x_k +\alpha_kS_k$ where $\alpha_k$ is the step size or the learning rate [@fletcher2013practical].  The step size is the key in the success of the optimization problem as inadequate step sizes might lead to a failure in convergence. These methods are known to be relative slow and can have convergence issues when problems are poorly conditioned [@trefethen1997numerical]. In order to solve some of those problems, line minimization or line search methods were proposed [@shi2004convergence] where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate $x_{1}$, a search direction $S_k$ for each iteration where the objective is to find the value of $\alpha_k \in \mathbb{R}$ that maximises $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ (this step is called the line search). Then move on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and repeat the process until the desire convergence is reached.
Modern development of line search methods focuses on improving the computation of the searching direction $S_{k}$, and on approximations of the step size $\alpha_k$ together with reducing the computational storing
burden of the intermediate computations while catering for practical optimisation problems. In the next section two line search methods for the optimisation used in the projection pursuit guided tour will be presented.

### 2020-12-10 paper/paper.tex
optim method section is good for proofreading

Visualisation is widely used in exploratory data analysis
\citep{tukey1977exploratory, unwin2015graphical, healy2018data, wilke2019fundamentals}.
Presenting information in graphics often unveils information that would
otherwise not be aware of and provides a more comprehensive
understanding of the problem at hand. The work presented in this paper
brings visualization tools into optimisation problems with an aim to
better understand the performance of the optimisers in practice. This
motivates our work of creating plots to diagnose optimisation algorithms
which in turn can be also used not only to understand optimization
performance but also to understand and compare the behaviour of
different algorithms.

@@

The motivation of this paper is the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational} which is an
exploratory data analysis tool that is defined to detect
\emph{interesting structures} or features in high-dimensional data
through a set of lower-dimensional projections that cover the entire
high dimensional space using interpolation methods called tours
\citep{cook2008grand}. The target of the optimisation is to identify the
most \emph{interesting} low-dimensional views of the data given by a
corresponding projection matrix. The most \emph{interesting} structures
are formally defined by a function of projections, called index function
which is optimized to uncover the most reveling structures in a high
dimensional space \citep{cook1993projection}.

@@

Optimization methods can be divided in various classes (NOTE: do we want
one more sentence on various class: global optim, convex optim,
stochastic optim etc?)
\citep{kelley1999iterative, boyd2004convex, fletcher2013practical}. In
this paper, we focused on methods that rely on the objective function
derivatives known as gradient methods, and those which are known as
derivative free methods where optimization does not rely on the gradient
or derivatives information.

In gradient descent methods the goal is to find the minimum of an
objective function while in gradient ascent methods the interest is in
the maximum. The minimum is searched for by taken steps proportional to
the negative gradient or of the function at the evaluation point
(analogously positive gradient for gradient ascent methods), that is
\(x_{k+1} = x_k -\alpha_kS_k\) where \(\alpha_k\) is the step size or
the learning rate \citep{fletcher2013practical}. The step size is the
key in the success of the optimization problem as inadequate step sizes
might lead to a failure in convergence. These methods are known to be
relative slow and can have convergence issues when problems are poorly
conditioned \citep{trefethen1997numerical}. In order to solve some of
those problems, line minimization or line search methods were proposed
\citep{shi2004convergence} where the step size is chosen in such a way
that convergence is ensured. In line search methods users are required
to provide an initial estimate \(x_{1}\), a search direction \(S_k\) for
each iteration where the objective is to find the value of
\(\alpha_k \in \mathbb{R}\) that minimizes \(f(x_k + \alpha_kS_k)\) with
respect to \(\alpha_k\) (this step is called the line search). Then move
on to the next point following \(x_{k+1} = x_k + \alpha_kS_k\) and
repeat the process until the desire convergence is reached. Modern
development of line search methods focuses on improving the computation
of the searching direction \(S_{k}\), and on approximations of the step
size \(\alpha_k\) together with reducing the computational storing
burden of the intermediate computations while catering for practical
optimisation problems. In the next section two line search methods for
the optimisation used in the projection pursuit guided tour will be
presented.

>>>>

Visualisation is widely used in exploratory data analysis
\citep{tukey1977exploratory, unwin2015graphical, healy2018data, wilke2019fundamentals}.
Presenting information in graphics often unveils information that would
otherwise not be aware of and provides a more comprehensive
understanding of the problem at hand. The work presented in this paper
brings visualization tools into optimisation problems with an aim to
better understand the performance of the optimisers in practice.

@@

The focus of this paper is on the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational} which is an
exploratory data analysis tool that is defined to detect
\emph{interesting structures} or features in high-dimensional data
through a set of lower-dimensional projections that cover the entire
high dimensional space using interpolation methods called tours
\citep{cook2008grand}. The target of the optimisation is to identify the
most \emph{interesting} low-dimensional views of the data given by a
corresponding projection matrix. The most \emph{interesting} structures
are formally defined by a function of projections, called index function
which is optimized to uncover the most reveling structures in a high
dimensional space \citep{cook1993projection}.

@@

Optimization methods can be divided in various classes, such as global
optimisation \citep{kelley1999iterative, fletcher2013practical}, convex
optimisation \citep{boyd2004convex}, stochastic optimisation
\citep{nocedal2006numerical} . In this paper, we focused on methods that
rely on the objective function derivatives known as gradient methods,
and those which are known as derivative free methods where optimization
does not rely on the gradient or derivatives information.

In gradient ascent methods the goal is to find the maximum of an
objective function while in gradient descent methods the interest is in
the minimum. The maximum is searched for by taken steps proportional to
the gradient of the function at the evaluation point that is
\(x_{k+1} = x_k +\alpha_kS_k\) where \(\alpha_k\) is the step size or
the learning rate \citep{fletcher2013practical}. The step size is the
key in the success of the optimization problem as inadequate step sizes
might lead to a failure in convergence. These methods are known to be
relative slow and can have convergence issues when problems are poorly
conditioned \citep{trefethen1997numerical}. In order to solve some of
those problems, line minimization or line search methods were proposed
\citep{shi2004convergence} where the step size is chosen in such a way
that convergence is ensured. In line search methods users are required
to provide an initial estimate \(x_{1}\), a search direction \(S_k\) for
each iteration where the objective is to find the value of
\(\alpha_k \in \mathbb{R}\) that maximises \(f(x_k + \alpha_kS_k)\) with
respect to \(\alpha_k\) (this step is called the line search). Then move
on to the next point following \(x_{k+1} = x_k + \alpha_kS_k\) and
repeat the process until the desire convergence is reached. Modern
development of line search methods focuses on improving the computation
of the searching direction \(S_{k}\), and on approximations of the step
size \(\alpha_k\) together with reducing the computational storing
burden of the intermediate computations while catering for practical
optimisation problems. In the next section two line search methods for
the optimisation used in the projection pursuit guided tour will be
presented.
